---
title: "Gerrymandering - Efficiency Gaps"
author: "Robbie Ha"
date: "2/6/2019"
output: 
  html_document

---

### Load packages

```{r load-packages, message=FALSE}
library(tidyverse) 
```

### 1976 Efficiency Gaps

I used data provided from the MIT election lab (https://electionlab.mit.edu/data) that provided house election data from 1976 to 2016. Utilizing the efficiency gap formula, I've calculated the efficiency gaps for this year, and will also be observing the EG's for 1996, and 2016.

```{r eg_1976, message=FALSE}
dn <- read_csv("1976-2016-house.csv")
eg_1976 <- dn %>% filter(year == 1976, party == "democrat" | party == "republican") %>% mutate(realDistrict = paste(state, district, sep="-")) %>% select(year, state, realDistrict, party, candidatevotes, totalvotes, state) %>% group_by(state) %>% summarise(republicanvotes = sum(candidatevotes[party=="republican"]),demvotes = sum(candidatevotes[party=="democrat"]), wastedvotes_rep = case_when(
  republicanvotes > demvotes ~ round(republicanvotes-((demvotes+republicanvotes)/2)-1, digit = 0),
  demvotes > republicanvotes ~ republicanvotes
),
wastedvotes_dem = case_when(
  republicanvotes > demvotes ~ demvotes,
  demvotes > republicanvotes ~ round(demvotes-((demvotes+republicanvotes)/2)-1, digit = 0)
),
  total_statevotes = demvotes+republicanvotes,
  eg = abs((wastedvotes_dem - wastedvotes_rep)/total_statevotes)) %>% select(state,eg)

glimpse(eg_1976)

```

The graph is ordered from the state with the highest efficiency gap to the lowest.

```{r eg_1976graph}

ggplot(eg_1976, mapping = aes(x = reorder(state,eg), y = eg, color = state)) +
  geom_point(alpha = 0.5) + coord_flip()

```

### 1996 Efficiency Gaps

```{r eg_1996, message=FALSE}
dn <- read_csv("1976-2016-house.csv")
eg_1996 <- dn %>% filter(year == 1996, party == "democrat" | party == "republican") %>% mutate(realDistrict = paste(state, district, sep="-")) %>% select(year, state, realDistrict, party, candidatevotes, totalvotes, state) %>% group_by(state) %>% summarise(republicanvotes = sum(candidatevotes[party=="republican"]),demvotes = sum(candidatevotes[party=="democrat"]), wastedvotes_rep = case_when(
  republicanvotes > demvotes ~ round(republicanvotes-((demvotes+republicanvotes)/2)-1, digit = 0),
  demvotes > republicanvotes ~ republicanvotes
),
wastedvotes_dem = case_when(
  republicanvotes > demvotes ~ demvotes,
  demvotes > republicanvotes ~ round(demvotes-((demvotes+republicanvotes)/2)-1, digit = 0)
),
  total_statevotes = demvotes+republicanvotes,
  eg = abs((wastedvotes_dem - wastedvotes_rep)/total_statevotes)) %>% select(state,eg)

glimpse(eg_1996)

```

The graph is ordered from the state with the highest efficiency gap to the lowest.

```{r eg_1996graph}

ggplot(eg_1996, mapping = aes(x = reorder(state,eg), y = eg, color = state)) +
  geom_point(alpha = 0.5) + coord_flip()

```

### 2016 Efficiency Gaps

```{r eg_2016, message=FALSE}
dn <- read_csv("1976-2016-house.csv")
eg_2016 <- dn %>% filter(year == 2016, party == "democrat" | party == "republican") %>% mutate(realDistrict = paste(state, district, sep="-")) %>% select(year, state, realDistrict, party, candidatevotes, totalvotes, state) %>% group_by(state) %>% summarise(republicanvotes = sum(candidatevotes[party=="republican"]),demvotes = sum(candidatevotes[party=="democrat"]), wastedvotes_rep = case_when(
  republicanvotes > demvotes ~ round(republicanvotes-((demvotes+republicanvotes)/2)-1, digit = 0),
  demvotes > republicanvotes ~ republicanvotes
),
wastedvotes_dem = case_when(
  republicanvotes > demvotes ~ demvotes,
  demvotes > republicanvotes ~ round(demvotes-((demvotes+republicanvotes)/2)-1, digit = 0)
),
  total_statevotes = demvotes+republicanvotes,
  eg = abs((wastedvotes_dem - wastedvotes_rep)/total_statevotes)) %>% select(state,eg)

glimpse(eg_2016)

```

The graph is ordered from the state with the highest efficiency gap to the lowest.

```{r eg_2016graph}

ggplot(eg_2016, mapping = aes(x = reorder(state,eg), y = eg, color = state)) +
  geom_point(alpha = 0.5) + coord_flip()

```

### Change from 1976 to 2016

The red dot signifies the efficiency gap in 2016, whereas the blue dot signifies the one for 1976.

```{r changegraph}

ggplot(eg_2016, mapping = aes(x = state, y = eg)) +
  geom_point(color="red") + 
  geom_point(eg_1976, mapping = aes(x = state, y = eg), color="blue") +coord_flip()


```
